\documentclass[12pt,a4paper]{article}

%--- Packages ---
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage[margin=2.5cm, headheight=15pt]{geometry}
\usepackage{amsmath,amssymb,amsfonts,amsthm}
\usepackage{mathtools}
\usepackage{bm}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{pgfplots}
\pgfplotsset{compat=1.18}
\usetikzlibrary{arrows.meta,shapes.geometric,shapes.misc,fit,positioning,
                calc,decorations.pathreplacing,backgrounds,matrix,shadows}
\usepackage{subcaption}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{array}
\usepackage{xcolor}
\usepackage{tcolorbox}
\tcbuselibrary{theorems,skins,breakable}
\usepackage{algorithm2e}
\usepackage{natbib}
\usepackage{hyperref}
\usepackage{cleveref}
\usepackage{setspace}
\usepackage{parskip}
\usepackage{titlesec}
\usepackage{fancyhdr}
\usepackage{caption}
\usepackage{mdframed}

%--- Colors ---
\definecolor{primaryblue}{RGB}{31,78,121}
\definecolor{accentorange}{RGB}{218,119,34}
\definecolor{lightgray}{RGB}{242,242,242}
\definecolor{darkgray}{RGB}{64,64,64}
\definecolor{mathgreen}{RGB}{0,128,0}
\definecolor{approachred}{RGB}{192,0,0}
\definecolor{approachblue}{RGB}{0,70,127}
\definecolor{approachpurple}{RGB}{112,48,160}
\definecolor{approachteal}{RGB}{0,112,112}
\definecolor{approachorange}{RGB}{180,90,0}

%--- tcolorbox styles ---
\tcbset{
  approachbox/.style={
    enhanced, breakable,
    colback=lightgray, colframe=primaryblue,
    fonttitle=\bfseries\color{white},
    coltitle=white,
    attach boxed title to top left={yshift=-2mm, xshift=4mm},
    boxed title style={colback=primaryblue, rounded corners=all},
    arc=4pt, outer arc=4pt,
    top=6pt, bottom=6pt
  }
}

%--- Page style ---
\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{\small\textit{Adaptive Dilation Factor Design for SSSDTCN}}
\fancyhead[R]{\small\thepage}
\fancyfoot[C]{\small\textit{Technical Research Plan -- Confidential}}
\renewcommand{\headrulewidth}{0.4pt}
\renewcommand{\footrulewidth}{0.4pt}

%--- Section formatting ---
\titleformat{\section}{\large\bfseries\color{primaryblue}}{{\thesection}}{1em}{}[\titlerule]
\titleformat{\subsection}{\normalsize\bfseries\color{darkgray}}{{\thesubsection}}{1em}{}
\titleformat{\subsubsection}{\normalsize\itshape\color{darkgray}}{{\thesubsubsection}}{1em}{}

%--- Math operators ---
\DeclareMathOperator{\FFT}{FFT}
\DeclareMathOperator{\TopK}{TopK}
\DeclareMathOperator{\softmax}{softmax}
\DeclareMathOperator{\sigmoid}{sigmoid}
\DeclareMathOperator{\floor}{floor}
\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\EE}{\mathbb{E}}
\newcommand{\XX}{\mathbf{X}}
\newcommand{\xv}{\mathbf{x}}
\newcommand{\yv}{\mathbf{y}}
\newcommand{\wv}{\mathbf{w}}
\newcommand{\hv}{\mathbf{h}}
\newcommand{\alphav}{\boldsymbol{\alpha}}
\newcommand{\thetav}{\boldsymbol{\theta}}

%--- Theorem environments ---
\newtheorem{definition}{Definition}
\newtheorem{proposition}{Proposition}
\newtheorem{remark}{Remark}

%=============================================================================
\begin{document}
%=============================================================================

%--- Title ---
\begin{titlepage}
  \begin{center}
    \vspace*{1.5cm}
    {\LARGE\bfseries\color{primaryblue}
      Adaptive Dilation Factor Design Strategies\\[6pt]
      for the Implicit Feature Extraction Module\\[6pt]
      in SSSDTCN-based Industrial Time Series Imputation}\\[1.5cm]

    {\large\color{darkgray} Technical Research Plan}\\[0.8cm]

    \rule{0.6\textwidth}{1pt}\\[0.8cm]

    {\normalsize
      \textit{Prepared for the SSSDTCN Project}\\[0.3cm]
      \textit{Branch: \texttt{claude/adaptive-dilation-factor-3y07x}}\\[0.3cm]
      \textit{Date: March 2026}}\\[1.5cm]

    \rule{0.6\textwidth}{0.5pt}\\[1cm]

    \begin{minipage}{0.85\textwidth}
      \begin{mdframed}[backgroundcolor=lightgray, linecolor=primaryblue,
                       linewidth=1.5pt, roundcorner=4pt]
        \textbf{Abstract.}
        The SSSDTCN model employs a dilated causal convolution stack as its
        implicit feature extraction module, with dilation rates fixed to an
        exponential schedule $\{1,2,4,8,16,32\}$ prior to training. While this
        hand-crafted design achieves competitive multi-scale temporal coverage,
        it treats all industrial time series uniformly and cannot adapt to the
        specific periodicity or correlation structure of a given dataset or
        input segment. This document surveys and formalises \textbf{five
        principled approaches} for making the dilation factors \emph{adaptive}:
        (I) learnable soft dilation parameters, (II) input-conditioned selective
        kernel fusion inspired by SKNet, (III) differentiable neural
        architecture search for discrete dilation rates (DARTS-based),
        (IV) frequency-spectral guided dilation derived from dominant period
        analysis, and (V) temporal self-attention guided progressive dilation.
        For each approach we state the mathematical formulation, describe the
        computational mechanism, identify the source literature, and discuss
        trade-offs. The goal is to provide a clear roadmap for selecting and
        implementing an adaptive dilation strategy within the existing SSSDTCN
        codebase.
      \end{mdframed}
    \end{minipage}

  \end{center}
\end{titlepage}

\tableofcontents
\newpage

%=============================================================================
\section{Introduction}
%=============================================================================

Industrial process control systems generate multivariate time series that
frequently suffer from missing observations due to sensor failures, network
interruptions, or scheduled maintenance. Accurate imputation of these missing
values is critical for downstream tasks such as process monitoring, fault
detection, and soft sensing. The SSSDTCN model~\citep{sssdtcn2024} addresses
this challenge by combining an \emph{implicit} feature extraction pathway---a
stack of dilated causal convolutions---with an \emph{explicit} pathway based on
Structured State Space Models (S4)~\citep{gu2021efficiently}, all embedded
within a denoising diffusion framework~\citep{ho2020denoising}.

The implicit pathway is responsible for capturing \emph{local multi-scale
temporal patterns}. Its current design uses a fixed exponential dilation
schedule $d_\ell = 2^{\ell-1}$ for layers $\ell=1,\ldots,L$, yielding dilation
rates $\{1,2,4,8,16,32\}$ for $L=6$. The total receptive field of a single
branch with kernel size $k$ and maximum dilation $d_{\max}$ is:
\begin{equation}
  \mathrm{RF} = 1 + (k-1)\sum_{\ell=1}^{L} d_\ell
              = 1 + (k-1)(2^L - 1).
  \label{eq:receptive_field}
\end{equation}
For $k=5$ and $L=6$ this gives $\mathrm{RF}=253$ time steps, offering broad
coverage. However, several limitations motivate adaptive designs:

\begin{enumerate}
  \item \textbf{Dataset heterogeneity.} The Debutanizer column has a dominant
        periodicity near 5--10 timesteps, while the SRU process exhibits
        long-range correlations up to 100+ timesteps. A uniform dilation schedule
        cannot be simultaneously optimal for both.

  \item \textbf{Missing-pattern dependence.} Under high missing-ratio regimes
        ($>60\%$), longer-range context becomes more valuable; under low
        missing ratios shorter-range dependencies suffice. Fixed dilations cannot
        rebalance in response to the observed missing pattern.

  \item \textbf{Non-stationarity.} Industrial processes can shift between
        operational regimes. Static dilation rates cannot track these changes.
\end{enumerate}

\begin{figure}[htbp]
  \centering
  \begin{tikzpicture}[>=Stealth, font=\small,
      box/.style={draw, rounded corners=3pt, minimum width=1.6cm,
                  minimum height=0.7cm, align=center, fill=white,
                  drop shadow},
      dilbox/.style={draw, rounded corners=3pt, minimum width=2cm,
                     minimum height=0.7cm, align=center},
      arr/.style={->, thick}
    ]
    % Input
    \node[box, fill=approachblue!20, text width=2.8cm] (inp) at (0,0)
          {Input\\$\xv \in \RR^{T \times C}$};

    % Dilation layers
    \foreach \d/\i/\col in {1/1/approachblue!10, 2/2/approachblue!20,
                             4/3/approachblue!30, 8/4/approachblue!40,
                             16/5/approachblue!50, 32/6/approachblue!60}{
      \node[dilbox, fill=\col, minimum width=2.4cm] (d\i) at (3.2, {2.5-(\i-1)*0.85})
            {$d_\ell=\d$\\\small dilation};
    }

    % Output
    \node[box, fill=accentorange!30, text width=2.8cm] (out) at (6.4, 0)
          {Implicit\\Features};

    % Arrows input -> layers
    \foreach \i in {1,...,6}{
      \draw[arr] (inp.east) -- (d\i.west);
    }

    % Arrows layers -> output
    \foreach \i in {1,...,6}{
      \draw[arr] (d\i.east) -- (out.west);
    }

    % Brace
    \draw[decorate,decoration={brace,amplitude=5pt,mirror}]
      ([xshift=2pt]d1.north east) -- ([xshift=2pt]d6.south east)
      node[midway, right=6pt, text width=2.2cm, align=center, font=\scriptsize]
      {Fixed schedule\\$d_\ell=2^{\ell-1}$};

    % Caption inside
    \node[above=3pt of d1, font=\scriptsize\itshape, text=darkgray]
          {Stack of dilated causal conv layers};
  \end{tikzpicture}
  \caption{Current fixed-dilation implicit module in SSSDTCN. All six dilation
           values are set before training and remain constant throughout.}
  \label{fig:current_arch}
\end{figure}

\Cref{fig:current_arch} illustrates the current design. The remainder of this
document presents five adaptive alternatives, each building on well-established
literature in deep learning and signal processing.

%=============================================================================
\section{Background and Notation}
%=============================================================================

\subsection{Dilated Causal Convolution}

Let $\xv \in \RR^{C \times T}$ denote a multivariate time series with $C$
channels and $T$ time steps. A 1-D dilated causal convolution with kernel
$\wv \in \RR^{C_\mathrm{out} \times C_\mathrm{in} \times k}$, kernel size $k$,
and dilation factor $d$ computes:
\begin{equation}
  (\xv \star_d \wv)[t] = \sum_{i=0}^{k-1} \wv[i] \cdot \xv[t - d \cdot i],
  \qquad t = 1,\ldots,T,
  \label{eq:dilated_conv}
\end{equation}
where \emph{causality} is enforced by zero-padding on the left with
$(k-1)\cdot d$ zeros and discarding right-side outputs. The effective receptive
field per layer is:
\begin{equation}
  \mathrm{RF}_\ell = 1 + (k-1)\cdot d_\ell.
  \label{eq:rf_single}
\end{equation}

\subsection{Current SSSDTCN Implicit Module}

The implicit module $f_\theta$ stacks $L$ dilated causal convolution layers
with learnable residual blending:
\begin{align}
  \hv^{(0)} &= \mathrm{Proj}(\xv), \label{eq:h0}\\
  \hv^{(\ell)} &= w_\ell \cdot \mathrm{Conv}_{d_\ell}(\hv^{(\ell-1)})
                + (1-w_\ell)\cdot \hv^{(\ell-1)},
                \quad \ell=1,\ldots,L, \label{eq:residual}\\
  f_\theta(\xv) &= \hv^{(L)}, \label{eq:output}
\end{align}
where $w_\ell \in [0,1]$ is a scalar learnable weight (initialised at 0.5)
and $d_\ell = 2^{\ell-1}$ is the \textbf{fixed} dilation factor. The adaptive
approaches below propose mechanisms to make $d_\ell$ (or the effective
multi-scale mixture) data-dependent.

\subsection{Terminology Summary}

\begin{center}
\begin{tabular}{ll}
  \toprule
  Symbol & Meaning \\
  \midrule
  $T$ & sequence length \\
  $C$ & number of channels / features \\
  $k$ & convolution kernel size \\
  $L$ & number of dilation layers \\
  $d_\ell$ & dilation factor of layer $\ell$ \\
  $D = \{d^{(1)},\ldots,d^{(K)}\}$ & candidate dilation set \\
  $\alphav$ & architecture / attention parameters \\
  $\thetav$ & model weight parameters \\
  \bottomrule
\end{tabular}
\end{center}

%=============================================================================
\section{Approach I: Learnable Soft Dilation Parameters}
%=============================================================================

\begin{tcolorbox}[approachbox, title={Approach I Overview}]
  \textbf{Key idea:} Replace the hand-crafted exponential schedule with
  unconstrained trainable scalars whose exponentiated values serve as
  continuous dilation factors, optimised end-to-end via back-propagation. \\[4pt]
  \textbf{Source:} Deformable Convolutional Networks~\citep{dai2017deformable};
  Temporal Deformable Convolution~\citep{chen2020temporal}.
\end{tcolorbox}

\subsection{Motivation}

Deformable convolutions~\citep{dai2017deformable} introduced the idea that
sampling offsets in a convolutional kernel need not be fixed on a regular grid
but can be learned from the data. Translated to the 1-D temporal setting, the
``offset'' concept corresponds directly to the dilation factor: rather than
sampling at equally-spaced positions $\{0, d, 2d, \ldots, (k-1)d\}$, we allow
the model to learn a data-driven effective stride.

\subsection{Mathematical Formulation}

For each layer $\ell$, introduce a scalar parameter $\lambda_\ell \in \RR$.
The effective dilation factor is:
\begin{equation}
  \boxed{d_\ell = \exp(\lambda_\ell),}
  \label{eq:soft_dilation}
\end{equation}
which guarantees $d_\ell > 0$ for all $\lambda_\ell$. During training,
gradients flow through the continuous relaxation of \cref{eq:dilated_conv}:
\begin{equation}
  (\xv \star_{d_\ell} \wv)[t]
    = \sum_{i=0}^{k-1} \wv[i] \cdot
      \underbrace{\mathrm{Interp}\!\left(\xv, t - e^{\lambda_\ell}\cdot i\right)
      }_{\text{linear interpolation for non-integer } d_\ell},
  \label{eq:soft_conv}
\end{equation}
where $\mathrm{Interp}(\xv, \tau) = (1-\{\tau\})\xv[\lfloor\tau\rfloor]
+ \{\tau\}\xv[\lceil\tau\rceil]$ is linear interpolation and $\{\cdot\}$
denotes the fractional part.

To prevent degenerate collapse (all dilations converging to the same value), a
diversity regulariser is added to the loss:
\begin{equation}
  \mathcal{L}_\mathrm{div} = -\beta \sum_{\ell < \ell'} \bigl|d_\ell - d_{\ell'}\bigr|,
  \label{eq:diversity_reg}
\end{equation}
with weight $\beta > 0$. The total training objective becomes:
\begin{equation}
  \mathcal{L} = \mathcal{L}_\mathrm{impute} + \mathcal{L}_\mathrm{div},
  \label{eq:total_loss_I}
\end{equation}
where $\mathcal{L}_\mathrm{impute}$ is the original masked MSE loss.

\subsection{Diagram}

\begin{figure}[htbp]
  \centering
  \begin{tikzpicture}[>=Stealth, font=\small,
      param/.style={draw, circle, fill=accentorange!40, minimum size=0.8cm},
      box/.style={draw, rounded corners=3pt, fill=approachblue!15,
                  minimum width=2.2cm, minimum height=0.7cm, align=center},
      arr/.style={->, thick}
    ]
    \node[box] (x) {$\hv^{(\ell-1)}$};

    % Learnable parameter
    \node[param, right=1.2cm of x] (lam) {$\lambda_\ell$};
    \node[below=0.1cm of lam, font=\scriptsize\itshape] {learnable};

    \node[box, right=1.0cm of lam] (exp) {$d_\ell = e^{\lambda_\ell}$};
    \node[box, right=1.0cm of exp] (conv)
          {$\xv\star_{d_\ell}\wv$\\[2pt]\scriptsize interp.};
    \node[box, right=1.0cm of conv] (out) {$\hv^{(\ell)}$};

    \draw[arr] (x) -- (lam);
    \draw[arr] (lam) -- (exp) node[midway,above,font=\scriptsize]{exp};
    \draw[arr] (exp) -- (conv) node[midway,above,font=\scriptsize]{dilation};
    \draw[arr] (conv) -- (out);
    \draw[arr] (x.north) to[out=60,in=120] node[midway,above,font=\scriptsize]{residual} (out.north);

    % Gradient arrow
    \draw[->, dashed, color=mathgreen, thick]
      (out.south) to[out=-90,in=-90]
      node[below,font=\scriptsize\color{mathgreen}]{$\nabla_{\lambda_\ell}\mathcal{L}$}
      (lam.south);
  \end{tikzpicture}
  \caption{Approach~I: learnable soft dilation. The parameter $\lambda_\ell$
           is optimised by back-propagation through the interpolated convolution.
           The green dashed arrow indicates the gradient flow.}
  \label{fig:approach1}
\end{figure}

\subsection{Advantages and Limitations}

\begin{itemize}
  \item[\textbf{+}] Minimal architectural change; only $L$ extra scalar parameters.
  \item[\textbf{+}] Fully differentiable; standard optimisers (Adam) apply.
  \item[\textbf{+}] Interpretable: the converged $d_\ell$ values directly reveal
        the dominant temporal scales of the dataset.
  \item[\textbf{--}] Dilation is still \emph{global} (shared across all time
        steps and samples); cannot adapt per-input.
  \item[\textbf{--}] Linear interpolation introduces approximation error;
        exact integer dilations would require a rounding operator that
        breaks the gradient.
  \item[\textbf{--}] The diversity regulariser requires tuning of $\beta$.
\end{itemize}

%=============================================================================
\section{Approach II: Input-Conditioned Selective Kernel Dilation}
%=============================================================================

\begin{tcolorbox}[approachbox, title={Approach II Overview}]
  \textbf{Key idea:} Run $K$ parallel dilated convolutions with distinct fixed
  candidate dilation rates, then use a squeeze-and-excitation attention
  mechanism to compute input-dependent soft weights for fusing their outputs.\\[4pt]
  \textbf{Source:} Selective Kernel Networks (SKNet)~\citep{li2019selective};
  Squeeze-and-Excitation Networks (SENet)~\citep{hu2018squeeze}.
\end{tcolorbox}

\subsection{Motivation}

SKNet~\citep{li2019selective} demonstrated that a network can dynamically select
which kernel size to emphasise for each input by routing through an
attention-based fusion. The same principle applies to dilation rates: given $K$
candidate dilations $D = \{d^{(1)},\ldots,d^{(K)}\}$, the model attends to the
scale most relevant for the current input segment.

\subsection{Mathematical Formulation}

\textbf{Step 1 --- Multi-scale feature extraction (Split):}
\begin{equation}
  \tilde{\hv}^{(j)} = \mathrm{BN}\!\left(\mathrm{Conv}_{d^{(j)}}(\hv^{(\ell-1)})\right),
  \quad j = 1,\ldots,K.
  \label{eq:sknet_split}
\end{equation}
Each branch uses dilation $d^{(j)}$ from the candidate set $D$.

\textbf{Step 2 --- Aggregation (Fuse):}
\begin{equation}
  \tilde{\mathbf{U}} = \sum_{j=1}^{K} \tilde{\hv}^{(j)},
  \qquad
  \mathbf{s} = \mathrm{GlobalAvgPool}(\tilde{\mathbf{U}}) \in \RR^{H},
  \label{eq:sknet_fuse}
\end{equation}
where $H$ is the hidden dimension (128 in SSSDTCN).

\textbf{Step 3 --- Compact representation (Squeeze):}
\begin{equation}
  \mathbf{z} = \mathrm{LayerNorm}\!\left(\mathrm{ReLU}(\mathbf{W}_1 \mathbf{s})\right)
  \in \RR^{r},
  \quad r = \max\!\left(H/4, 16\right),
  \label{eq:sknet_squeeze}
\end{equation}
with $\mathbf{W}_1 \in \RR^{r \times H}$.

\textbf{Step 4 --- Scale-aware attention (Select):}
\begin{equation}
  \boldsymbol{\alpha}^{(j)} = \frac{\exp(\mathbf{W}_2^{(j)}\mathbf{z})}
                                    {\sum_{j'=1}^K \exp(\mathbf{W}_2^{(j')}\mathbf{z})},
  \quad \mathbf{W}_2^{(j)} \in \RR^{H \times r},
  \label{eq:sknet_select}
\end{equation}
producing a channel-wise soft selection vector $\boldsymbol{\alpha}^{(j)} \in [0,1]^H$
for each branch, with $\sum_{j=1}^K \boldsymbol{\alpha}^{(j)} = \mathbf{1}$.

\textbf{Step 5 --- Weighted recombination:}
\begin{equation}
  \boxed{
  \hv^{(\ell)} = \sum_{j=1}^{K} \boldsymbol{\alpha}^{(j)} \odot \tilde{\hv}^{(j)},
  }
  \label{eq:sknet_output}
\end{equation}
where $\odot$ is element-wise multiplication with broadcasting over the time
axis.

The \emph{effective adaptive dilation} is implicitly encoded in the attention
weights:
\begin{equation}
  d_\ell^{\mathrm{eff}}(t) = \sum_{j=1}^{K} \bar{\alpha}^{(j)}(t) \cdot d^{(j)},
  \label{eq:effective_dilation}
\end{equation}
where $\bar{\alpha}^{(j)}(t)$ denotes the mean of $\boldsymbol{\alpha}^{(j)}$
over channels at time $t$. This provides an interpretable scalar measure of the
dominant operating scale.

\subsection{Diagram}

\begin{figure}[htbp]
  \centering
  \begin{tikzpicture}[>=Stealth, font=\small,
      box/.style={draw, rounded corners=3pt, fill=approachblue!15,
                  minimum width=2.0cm, minimum height=0.6cm, align=center},
      attbox/.style={draw, rounded corners=3pt, fill=accentorange!20,
                     minimum width=2.0cm, minimum height=0.6cm, align=center},
      arr/.style={->, thick}
    ]
    \node[box] (h) {$\hv^{(\ell-1)}$};

    % Parallel branches
    \node[box, fill=approachblue!10, right=1.5cm of h, yshift=1.2cm]
      (b1) {$d^{(1)}=1$};
    \node[box, fill=approachblue!20, right=1.5cm of h, yshift=0.4cm]
      (b2) {$d^{(2)}=2$};
    \node[box, fill=approachblue!30, right=1.5cm of h, yshift=-0.4cm]
      (b3) {$d^{(3)}=4$};
    \node[box, fill=approachblue!40, right=1.5cm of h, yshift=-1.2cm]
      (b4) {$d^{(4)}=8$};

    \draw[arr] (h) -- (b1.west);
    \draw[arr] (h) -- (b2.west);
    \draw[arr] (h) -- (b3.west);
    \draw[arr] (h) -- (b4.west);

    % Aggregation
    \node[attbox, right=1.2cm of b2] (agg) {Aggregate\\$\tilde{\mathbf{U}}$};
    \draw[arr] (b1.east) -- (agg.north west);
    \draw[arr] (b2.east) -- (agg.west);
    \draw[arr] (b3.east) -- (agg.west);
    \draw[arr] (b4.east) -- (agg.south west);

    % Squeeze-Excite
    \node[attbox, right=0.8cm of agg] (se) {GAP\\$\rightarrow$ FC\\$\rightarrow$ Attn};
    \draw[arr] (agg) -- (se);

    % Output
    \node[box, fill=accentorange!30, right=0.8cm of se] (out)
          {$\hv^{(\ell)}$\\[2pt]\scriptsize weighted\\sum};
    \draw[arr] (se) -- node[above,font=\scriptsize]{$\boldsymbol{\alpha}^{(j)}$} (out);
    \draw[arr] (b1.east) to[out=0, in=80] (out.north);
    \draw[arr] (b4.east) to[out=0, in=-80] (out.south);

    \node[font=\scriptsize\itshape, below=1.8cm of agg, text=darkgray]
      {Input-conditioned recombination};
  \end{tikzpicture}
  \caption{Approach~II: Selective Kernel dilation. Four candidate dilations run
           in parallel; the squeeze-and-excitation block produces per-channel
           soft attention weights that adaptively recombine the outputs.}
  \label{fig:approach2}
\end{figure}

\subsection{Advantages and Limitations}

\begin{itemize}
  \item[\textbf{+}] Fully input-adaptive: different time steps and samples can
        produce different effective dilations.
  \item[\textbf{+}] Channel-wise attention provides fine-grained selection;
        each feature dimension can attend to a different scale.
  \item[\textbf{+}] Well-grounded in the SKNet and SENet literature with strong
        empirical results in vision; the translation to temporal data is
        natural.
  \item[\textbf{--}] Computational cost scales as $\mathcal{O}(K)$; memory usage
        grows with the number of parallel branches.
  \item[\textbf{--}] The candidate set $D$ must still be pre-specified (though
        the weights between candidates are learned).
\end{itemize}

%=============================================================================
\section{Approach III: Differentiable Architecture Search for Dilation Rates}
%=============================================================================

\begin{tcolorbox}[approachbox, title={Approach III Overview}]
  \textbf{Key idea:} Treat dilation-rate selection as a discrete architecture
  decision and relax it into a continuous optimisation using the
  DARTS~\citep{liu2019darts} bi-level framework. After search, the
  architecture is discretised to the top-1 rate per layer.\\[4pt]
  \textbf{Source:} DARTS~\citep{liu2019darts}; P-DARTS~\citep{chen2019progressive};
  Auto-TCN~\citep{autotcn2022}.
\end{tcolorbox}

\subsection{Motivation}

Neural Architecture Search (NAS) has demonstrated that the optimal
computational operation at each network layer can be discovered automatically.
DARTS~\citep{liu2019darts} introduced a continuous relaxation over a discrete
search space via a \emph{mixed operation}, making the search differentiable.
For our setting, the search space at layer $\ell$ is the set of $K$ candidate
dilated convolutions $\mathcal{O}_\ell = \{\mathrm{Conv}_{d^{(j)}}: j=1,\ldots,K\}$.

\subsection{Mathematical Formulation}

\textbf{Mixed operation:}
During the \emph{search phase}, each layer computes a weighted mixture over all
candidate operations:
\begin{equation}
  \bar{o}^{(\ell)}(\hv) = \sum_{j=1}^{K}
    \underbrace{\frac{\exp(\alpha_\ell^{(j)})}{\sum_{j'=1}^{K}\exp(\alpha_\ell^{(j')})}
    }_{\pi_\ell^{(j)} \triangleq \text{arch. weight}}
    \mathrm{Conv}_{d^{(j)}}(\hv),
  \label{eq:darts_mixed}
\end{equation}
where $\boldsymbol{\alpha}_\ell = (\alpha_\ell^{(1)},\ldots,\alpha_\ell^{(K)}) \in \RR^K$
are the \textbf{architecture parameters} (one vector per layer, shared across
samples).

\textbf{Bi-level optimisation:}
Let $\thetav$ be the convolution weights and $\alphav$ be the architecture
parameters. The search solves:
\begin{align}
  \min_{\alphav} \quad & \mathcal{L}_\mathrm{val}\!\left(\thetav^*(\alphav), \alphav\right)
  \label{eq:darts_upper}\\
  \text{s.t.} \quad & \thetav^*(\alphav) =
    \argmin_{\thetav} \mathcal{L}_\mathrm{train}(\thetav, \alphav).
  \label{eq:darts_lower}
\end{align}
In practice, first-order DARTS approximates this by alternating gradient steps:
\begin{equation}
  \thetav \leftarrow \thetav - \eta_\theta \nabla_{\thetav}
    \mathcal{L}_\mathrm{train}(\thetav, \alphav), \qquad
  \alphav \leftarrow \alphav - \eta_\alpha \nabla_{\alphav}
    \mathcal{L}_\mathrm{val}(\thetav, \alphav).
  \label{eq:darts_update}
\end{equation}

\textbf{Discretisation:}
After the search phase, the dilation at each layer is fixed to:
\begin{equation}
  d_\ell^* = d^{(\hat{j}_\ell)}, \qquad
  \hat{j}_\ell = \argmax_{j} \alpha_\ell^{(j)}.
  \label{eq:darts_discrete}
\end{equation}
The selected architecture is then \emph{retrained from scratch} with the
discrete dilations.

\subsection{Search Space Design for SSSDTCN}

We recommend the following DARTS setup:
\begin{center}
\begin{tabular}{ll}
  \toprule
  Hyper-parameter & Value \\
  \midrule
  Candidate dilations $D$ & $\{1, 2, 3, 4, 6, 8, 12, 16, 24, 32\}$ \\
  Layers $L$ & 6 (one architecture param vector per layer) \\
  Search epochs & 30 (using 20\% validation split) \\
  $\eta_\theta$ (weight LR) & $10^{-3}$ \\
  $\eta_\alpha$ (arch LR)  & $3\times10^{-4}$ \\
  Arch. weight init & $\alpha_\ell^{(j)} = 0$ (uniform mixture) \\
  \bottomrule
\end{tabular}
\end{center}

\subsection{Diagram}

\begin{figure}[htbp]
  \centering
  \begin{tikzpicture}[>=Stealth, font=\small,
      snode/.style={draw, circle, fill=approachpurple!20,
                    minimum size=0.9cm, align=center},
      ebox/.style={draw, rounded corners=3pt, fill=approachpurple!10,
                   minimum width=1.8cm, minimum height=0.6cm, align=center},
      arr/.style={->, thick},
      mixed/.style={draw, rounded corners=3pt, fill=accentorange!20,
                    minimum width=3.0cm, minimum height=1.6cm, align=center}
    ]
    \node[snode] (n0) at (0,0) {$\hv^{(0)}$};
    \node[snode] (n1) at (6,0) {$\hv^{(1)}$};

    % Mixed operation box
    \node[mixed] (mix) at (3,0)
      {$\bar{o}(\hv) = \sum_j \pi^{(j)}_\ell \mathrm{Conv}_{d^{(j)}}$\\
       \vspace{2pt}\scriptsize $d^{(j)}\in\{1,2,3,4,6,8,12,16,24,32\}$};

    \draw[arr] (n0) -- (mix);
    \draw[arr] (mix) -- (n1);

    % Architecture parameters
    \node[ebox, fill=approachpurple!30, above=1cm of mix] (alpha)
          {$\boldsymbol{\alpha}_\ell \in \RR^{10}$\\[2pt]\scriptsize arch. params};
    \draw[->, dashed, color=approachpurple, thick] (alpha.south) -- (mix.north);

    % Gradient arrows
    \draw[->, dotted, color=mathgreen, bend right=30, thick]
      (n1.north) to node[midway,above,font=\scriptsize\color{mathgreen}]
      {$\nabla_{\alphav} \mathcal{L}_\mathrm{val}$} (alpha.east);
    \draw[->, dotted, color=approachblue, bend left=30, thick]
      (n1.south) to node[midway,below,font=\scriptsize\color{approachblue}]
      {$\nabla_{\thetav} \mathcal{L}_\mathrm{train}$} (n0.south);

    % Discretise arrow
    \node[ebox, fill=accentorange!30, right=1.2cm of n1] (disc)
          {Discretise\\$d^* = d^{(\hat{j})}$};
    \draw[->, thick, color=approachred] (n1) -- (disc)
      node[midway, above, font=\scriptsize]{after search};
  \end{tikzpicture}
  \caption{Approach~III: DARTS-based architecture search. During the search
           phase, a mixture of all candidate dilations is used. After search,
           the dilation with the highest architecture weight is selected.}
  \label{fig:approach3}
\end{figure}

\subsection{Advantages and Limitations}

\begin{itemize}
  \item[\textbf{+}] Searches over a much richer dilation space than the
        original exponential schedule.
  \item[\textbf{+}] Architecture parameters are shared globally, so the
        selected dilations reflect dataset-level (not just sample-level)
        optimal scales.
  \item[\textbf{+}] After discretisation, inference has \emph{the same cost}
        as a fixed-dilation model.
  \item[\textbf{--}] Requires two optimisation phases (search + retrain).
  \item[\textbf{--}] Bi-level optimisation is sensitive to hyper-parameters;
        may require $\alpha$-regularisation to avoid collapsing to one dilation.
  \item[\textbf{--}] DARTS can suffer from ``discretisation gap'' --- the
        mixed-operation performance during search may not faithfully predict
        single-operation performance after discretisation.
\end{itemize}

%=============================================================================
\section{Approach IV: Frequency-Spectral Guided Adaptive Dilation}
%=============================================================================

\begin{tcolorbox}[approachbox, title={Approach IV Overview}]
  \textbf{Key idea:} Analyse the frequency spectrum of the input sequence to
  identify its dominant periodic components, then set the dilation rates
  \emph{proportional to those periods} so each layer ``resonates'' with one
  dominant temporal scale.\\[4pt]
  \textbf{Source:} TimesNet~\citep{wu2023timesnet}; FEDformer~\citep{zhou2022fedformer};
  Autoformer~\citep{wu2021autoformer}.
\end{tcolorbox}

\subsection{Motivation}

TimesNet~\citep{wu2023timesnet} demonstrated that industrial and weather time
series contain strong periodic structure. If the dominant periods of the series
are known (or estimated on-line), the dilation rates can be calibrated to match
them exactly, ensuring that the dilated convolution's sampling pattern aligns
with the natural temporal rhythm of the process.

\subsection{Mathematical Formulation}

\textbf{Step 1 --- Discrete Fourier Transform:}
Given the input segment $\xv \in \RR^{C \times T}$, compute the average
spectral magnitude across channels:
\begin{equation}
  A_f = \frac{1}{C}\sum_{c=1}^C \left|\FFT(\xv[c,:])[f]\right|,
  \quad f = 0,1,\ldots,\lfloor T/2 \rfloor,
  \label{eq:fft_mag}
\end{equation}
where $\FFT(\cdot)$ denotes the Fast Fourier Transform.

\textbf{Step 2 --- Dominant period extraction:}
\begin{equation}
  \{f_1,\ldots,f_L\} = \TopK(\{A_f\}_{f=1}^{\lfloor T/2\rfloor}, K=L),
  \label{eq:topk_freq}
\end{equation}
excluding $f=0$ (DC component). The corresponding periods are:
\begin{equation}
  P_\ell = \left\lfloor \frac{T}{f_\ell} \right\rceil,
  \quad \ell = 1,\ldots,L,
  \label{eq:period}
\end{equation}
where $\lfloor\cdot\rceil$ denotes rounding to the nearest integer.

\textbf{Step 3 --- Period-to-dilation mapping:}
The dilation for layer $\ell$ is set so that the receptive field of a
\emph{single} kernel step spans exactly one period:
\begin{equation}
  \boxed{
  d_\ell = \max\!\left(1,\, \left\lfloor \frac{P_\ell}{k - 1} \right\rceil\right),
  }
  \label{eq:freq_dilation}
\end{equation}
ensuring $\mathrm{RF}_\ell = 1 + (k-1)d_\ell \approx P_\ell + 1$.
The layers are sorted in ascending order of $P_\ell$ to preserve the
coarse-to-fine stack ordering.

\textbf{Step 4 --- Normalisation and deduplication:}
If two computed dilations coincide, the less dominant one is perturbed by
$+1$ to maintain layer diversity:
\begin{equation}
  \text{if } d_\ell = d_{\ell'}, \ell \neq \ell': \quad
  d_{\ell'} \leftarrow d_{\ell'} + 1.
  \label{eq:dedup}
\end{equation}

\textbf{Online update (optional):}
For non-stationary processes, the dilation rates can be recomputed at the
start of each mini-batch using a moving-average periodogram:
\begin{equation}
  \hat{A}_f^{(t)} = (1-\gamma)\hat{A}_f^{(t-1)} + \gamma A_f^{(t)},
  \label{eq:ema_spectrum}
\end{equation}
with decay factor $\gamma \in (0,1)$. This introduces zero additional
learnable parameters.

\subsection{Diagram}

\begin{figure}[htbp]
  \centering
  \begin{tikzpicture}[>=Stealth, font=\small,
      box/.style={draw, rounded corners=3pt, minimum width=2.0cm,
                  minimum height=0.7cm, align=center, fill=approachteal!15},
      arr/.style={->, thick}
    ]
    \node[box] (inp) {Input $\xv$};

    \node[box, fill=approachteal!25, right=0.9cm of inp] (fft)
          {FFT\\$|F_f|$};
    \node[box, fill=approachteal!35, right=0.9cm of fft] (topk)
          {Top-$L$\\freqs};
    \node[box, fill=approachteal!45, right=0.9cm of topk] (period)
          {Periods\\$P_\ell = T/f_\ell$};
    \node[box, fill=accentorange!40, right=0.9cm of period] (dil)
          {Dilations\\$d_\ell=\lfloor P_\ell/(k-1)\rceil$};

    \draw[arr] (inp) -- (fft);
    \draw[arr] (fft) -- (topk);
    \draw[arr] (topk) -- (period);
    \draw[arr] (period) -- (dil);

    % Spectrum plot
    \begin{scope}[shift={(1.2,-1.8)}]
      \begin{axis}[
          width=3.5cm, height=2.0cm,
          xtick=\empty, ytick=\empty,
          axis lines=left,
          xlabel={\scriptsize freq $f$},
          ylabel={\scriptsize $|A_f|$},
          x label style={at={(axis description cs:0.5,-0.15)},anchor=north,font=\tiny},
          y label style={at={(axis description cs:-0.25,0.5)},anchor=south,font=\tiny},
        ]
        \addplot[fill=approachteal!30, draw=approachteal!70, ybar, bar width=1.5pt,
                 samples at={1,2,3,4,5,6,7,8,9,10,11,12}]
          {0.1+0.7*exp(-0.4*(x-3)^2)+0.5*exp(-0.3*(x-7)^2)+0.3*exp(-0.5*(x-11)^2)};
        \addplot[mark=*, mark size=2pt, color=approachred]
          coordinates{(3,0.8) (7,0.6) (11,0.4)};
      \end{axis}
    \end{scope}

    \node[font=\scriptsize\itshape, below=0.5cm of topk, text=darkgray,
          xshift=-0.8cm]
      {Dominant peaks $\rightarrow$ periods $\rightarrow$ dilations};
  \end{tikzpicture}
  \caption{Approach~IV: frequency-spectral guided dilation. The dominant
           spectral frequencies are extracted via FFT, converted to periods,
           and mapped to per-layer dilation rates. The spectrum inset shows
           three dominant peaks (red dots) at frequencies $f=3,7,11$.}
  \label{fig:approach4}
\end{figure}

\subsection{Worked Example for the Debutanizer Dataset}

The Debutanizer column has a dominant oscillation at approximately 6-step
cycles~\citep{sssdtcn2024}. For a sequence of $T=60$ and $k=5$:
\begin{align}
  f_1 &= 10, \quad P_1 = 60/10 = 6 \text{ steps}
      \quad \Rightarrow \quad d_1 = \max(1, \lfloor 6/4 \rfloor) = 1,\\
  f_2 &= 5,  \quad P_2 = 60/5  = 12 \text{ steps}
      \quad \Rightarrow \quad d_2 = \lfloor 12/4 \rfloor = 3,\\
  f_3 &= 3,  \quad P_3 = 60/3  = 20 \text{ steps}
      \quad \Rightarrow \quad d_3 = \lfloor 20/4 \rfloor = 5,
\end{align}
which differs from the fixed schedule $\{1,2,4\}$ and is calibrated to the
physical periodicity of the process.

\subsection{Advantages and Limitations}

\begin{itemize}
  \item[\textbf{+}] Zero learnable parameters for the dilation-selection step;
        no additional optimisation required.
  \item[\textbf{+}] Directly interpretable: dilations correspond to physically
        meaningful process periods.
  \item[\textbf{+}] Can be updated on-line for non-stationary processes at
        negligible computational cost.
  \item[\textbf{--}] Assumes periodicity is the primary temporal structure; may
        not generalise to trend-dominated or aperiodic series.
  \item[\textbf{--}] Sensitive to noise in the spectrum; a low-pass smoothing
        step may be needed before peak detection.
  \item[\textbf{--}] Dilation rates are discrete-valued and may not exactly
        match computed periods.
\end{itemize}

%=============================================================================
\section{Approach V: Temporal Self-Attention Guided Progressive Dilation}
%=============================================================================

\begin{tcolorbox}[approachbox, title={Approach V Overview}]
  \textbf{Key idea:} Insert a lightweight temporal self-attention module before
  the dilation stack to estimate a \emph{context-relevance score} for each
  time offset. These scores modulate the convolution kernels dynamically,
  producing an attention-weighted effective dilation that is different for
  each position.\\[4pt]
  \textbf{Source:} Temporal Convolutional Attention Network (TCAN)~\citep{tcan2020};
  Dynamic Convolution~\citep{chen2020dynamic};
  Non-local Means~\citep{wang2018nonlocal}.
\end{tcolorbox}

\subsection{Motivation}

While Approaches I--IV adapt the dilation either globally (I, III) or through
parallel branches (II, IV), they do not modify the \emph{kernel itself} based
on position-level context. Dynamic Convolution~\citep{chen2020dynamic} showed
that conditioning the convolution kernel on the current input can significantly
improve representation capacity. Here, we extend this idea to the temporal
dilation setting.

\subsection{Mathematical Formulation}

\textbf{Step 1 --- Temporal attention scores:}
Given the hidden sequence $\hv^{(\ell-1)} \in \RR^{H \times T}$, compute a
scalar attention score for each possible time offset $\delta = 0,1,\ldots,T_\mathrm{max}$:
\begin{equation}
  e_{t,\delta} = \frac{\mathbf{q}_t^\top \mathbf{k}_{t-\delta}}{\sqrt{d_k}},
  \qquad
  a_{t,\delta} = \frac{\exp(e_{t,\delta})}
                      {\sum_{\delta'=0}^{T_\mathrm{max}}\exp(e_{t,\delta'})},
  \label{eq:attn_score}
\end{equation}
where $\mathbf{q}_t = \mathbf{W}_Q\hv^{(\ell-1)}[:,t]$ and
$\mathbf{k}_t = \mathbf{W}_K\hv^{(\ell-1)}[:,t]$ are query/key projections,
and $d_k$ is the key dimension.

\textbf{Step 2 --- Effective dilation estimation:}
The expected offset (effective dilation) at position $t$ is:
\begin{equation}
  \bar{d}_t = \sum_{\delta=0}^{T_\mathrm{max}} \delta \cdot a_{t,\delta}.
  \label{eq:expected_dilation}
\end{equation}
The \emph{position-level effective dilation} $\bar{d}_t$ naturally adapts:
near boundaries or in regions with long-range context, attention concentrates
on large offsets (large $\bar{d}_t$); near high-frequency transitions, it
concentrates on small offsets.

\textbf{Step 3 --- Attention-guided dynamic convolution:}
For layer $\ell$ with base dilation $d_\ell^{(0)}$ (the original fixed
schedule), the modulated output is:
\begin{equation}
  \boxed{
  \hv^{(\ell)}[:,t] = \sum_{i=0}^{k-1}
    \underbrace{g_{t,i}}_{\text{gate}} \cdot
    \wv_\ell[:,i] \cdot
    \hv^{(\ell-1)}\!\left[:,t - \mathrm{round}\!\left(a_{t,i}\cdot d_\ell^{(0)}\right)\right],
  }
  \label{eq:dynamic_conv}
\end{equation}
where the gate $g_{t,i} = \sigmoid(\mathbf{W}_g\hv^{(\ell-1)}[:,t] + b_g) \in (0,1)$
controls the contribution of each kernel tap. A simplified version uses the
aggregate score:
\begin{equation}
  \tilde{d}_{t,\ell} = d_\ell^{(0)} \cdot \sigmoid\!\left(\mathbf{W}_s
    \hv^{(\ell-1)}[:,t] + b_s\right) \cdot 2,
  \label{eq:dynamic_dilation_simple}
\end{equation}
which allows the dilation to vary within $[0, 2d_\ell^{(0)}]$ depending on
local context, providing continuous input-dependent adaptation.

\textbf{Computational complexity:}
The attention step adds $\mathcal{O}(T_\mathrm{max} \cdot T \cdot d_k)$ per
layer. Using $T_\mathrm{max} = d_\ell^{(0)} \cdot k$ (only attend within the
original receptive field) keeps the overhead proportional to the base
convolution cost.

\subsection{Diagram}

\begin{figure}[htbp]
  \centering
  \begin{tikzpicture}[>=Stealth, font=\small,
      box/.style={draw, rounded corners=3pt, fill=approachorange!15,
                  minimum width=2.2cm, minimum height=0.7cm, align=center},
      gatebox/.style={draw, rounded corners=3pt, fill=accentorange!25,
                      minimum width=2.0cm, minimum height=0.7cm, align=center},
      arr/.style={->, thick}
    ]
    \node[box] (h0) {$\hv^{(\ell-1)}$};

    % Q, K projections
    \node[box, fill=approachorange!20, right=0.8cm of h0, yshift=0.9cm]
      (q) {$\mathbf{q}_t$};
    \node[box, fill=approachorange!20, right=0.8cm of h0, yshift=-0.9cm]
      (k) {$\mathbf{k}_t$};

    \draw[arr] (h0) -- (q);
    \draw[arr] (h0) -- (k);

    % Attention scores
    \node[gatebox, right=0.9cm of q, yshift=-0.9cm] (attn)
          {Attention\\$a_{t,\delta}$};
    \draw[arr] (q) -- (attn);
    \draw[arr] (k) -- (attn);

    % Gate
    \node[gatebox, fill=accentorange!40, right=1.0cm of attn] (gate)
          {Gate\\$g_{t,i}$};
    \draw[arr] (attn) -- node[above,font=\scriptsize]{$\bar{d}_t$} (gate);
    \draw[arr] (h0.east) to[out=0, in=-120] (gate.south);

    % Output
    \node[box, fill=approachorange!35, right=1.0cm of gate] (out)
          {$\hv^{(\ell)}$};
    \draw[arr] (gate) -- (out);

    % Base dilation annotation
    \node[font=\scriptsize, below=0.2cm of attn, text=darkgray]
          {base $d_\ell^{(0)}$};
    \node[font=\scriptsize, below=0.2cm of gate, text=darkgray]
          {adapted $\tilde{d}_{t,\ell}$};
  \end{tikzpicture}
  \caption{Approach~V: temporal self-attention guided dynamic dilation. The
           attention scores $a_{t,\delta}$ estimate the expected offset at each
           position $t$, producing a position-level effective dilation that
           modulates the base convolution.}
  \label{fig:approach5}
\end{figure}

\subsection{Advantages and Limitations}

\begin{itemize}
  \item[\textbf{+}] Most expressive of the five approaches: the dilation is
        position-dependent and sample-dependent.
  \item[\textbf{+}] Naturally handles non-stationary series by adapting
        per-position.
  \item[\textbf{+}] Backed by extensive literature on dynamic and deformable
        convolutions.
  \item[\textbf{--}] Higher computational cost than the other approaches.
  \item[\textbf{--}] The attention mechanism introduces additional parameters
        ($\mathbf{W}_Q$, $\mathbf{W}_K$, $\mathbf{W}_g$) per layer.
  \item[\textbf{--}] Non-integer dilations require interpolation; the resulting
        gradient landscape can be more complex to optimise.
\end{itemize}

%=============================================================================
\section{Comparative Analysis}
%=============================================================================

\subsection{Quantitative Comparison Table}

\begin{table}[htbp]
  \centering
  \caption{Comparison of the five adaptive dilation approaches along key
           design dimensions. ``Global'' means the same dilation applies to all
           input positions; ``Per-sample'' means different inputs can produce
           different dilations; ``Per-position'' is the most fine-grained.}
  \label{tab:comparison}
  \setlength{\tabcolsep}{5pt}
  \begin{tabular}{@{}lcccccl@{}}
    \toprule
    \multirow{2}{*}{\textbf{Approach}} &
    \multirow{2}{*}{\textbf{Extra params}} &
    \textbf{Adapt.} &
    \multirow{2}{*}{\textbf{Search?}} &
    \multirow{2}{*}{\textbf{Inference cost}} &
    \textbf{Impl.} &
    \multirow{2}{*}{\textbf{Key reference}} \\
    & & \textbf{level} & & & \textbf{complexity} & \\
    \midrule
    Fixed (baseline) & 0 & --- & No & $\mathcal{O}(LC)$ & Very low &---\\
    \midrule
    I.\ Soft dilation  & $L$  scalars & Global & No & $\mathcal{O}(LC)$ & Low &\cite{dai2017deformable}\\
    II.\ SKNet fusion & $2r\cdot H\cdot K$ & Per-sample & No & $\mathcal{O}(KLC)$ & Medium &\cite{li2019selective}\\
    III.\ DARTS & $LK$ (search only) & Global & Yes (2-phase) & $\mathcal{O}(LC)$ & High &\cite{liu2019darts}\\
    IV.\ Spectral & 0 & Global & No & $\mathcal{O}(LC)$ & Low &\cite{wu2023timesnet}\\
    V.\ Attn-guided & $3Hd_k\cdot L$ & Per-position & No & $\mathcal{O}(LC\cdot T)$ & High &\cite{chen2020dynamic}\\
    \bottomrule
  \end{tabular}
\end{table}

\subsection{Receptive Field Coverage}

\Cref{fig:rf_compare} compares the total receptive field $\mathrm{RF}$ achievable
by each approach. Approaches II and V can in principle cover \emph{the full
sequence} if their attention assigns weight to the maximum offset; Approaches
I, III, and IV are bounded by the largest learned or searched dilation.

\begin{figure}[htbp]
  \centering
  \begin{tikzpicture}
    \begin{axis}[
        width=12cm, height=5cm,
        ybar=0.3cm,
        bar width=0.6cm,
        xtick={1,2,3,4,5,6},
        xticklabels={Fixed, Soft~(I), SKNet~(II), DARTS~(III), Spectral~(IV), Attn~(V)},
        xticklabel style={font=\small, rotate=15, anchor=east},
        ylabel={Max.\ receptive field (timesteps)},
        ylabel style={font=\small},
        ymin=0, ymax=320,
        ymajorgrids=true,
        legend pos=north west,
        legend style={font=\small},
        axis lines=left,
        enlarge x limits=0.1,
      ]
      \addplot[fill=approachblue!50] coordinates {
        (1, 253) (2, 253) (3, 253) (4, 253) (5, 253) (6, 253)};
      \addlegendentry{Fixed-schedule RF}
      \addplot[fill=accentorange!70] coordinates {
        (1, 253) (2, 300) (3, 253) (4, 300) (5, 240) (6, 300)};
      \addlegendentry{Adaptive (potential)}
    \end{axis}
  \end{tikzpicture}
  \caption{Illustrative comparison of achievable receptive field for the fixed
           baseline versus the five adaptive approaches. Approaches I, III, IV
           learn or compute optimal rates but are bounded; II and V can
           adaptively increase effective coverage (orange bars represent
           potential under favourable conditions).}
  \label{fig:rf_compare}
\end{figure}

\subsection{Recommended Strategy for SSSDTCN}

Based on the analysis above, we recommend a two-stage strategy:

\begin{enumerate}
  \item \textbf{Stage 1 --- Offline period analysis (Approach IV):}
        Run the spectral analysis once on the training set to identify the top
        $L$ dominant periods. Use the resulting $\{d_\ell^*\}$ as the initial
        dilation schedule. This costs nothing at training time and provides a
        physics-informed initialisation.

  \item \textbf{Stage 2 --- Fine-tuning with soft dilation (Approach I):}
        Initialise $\lambda_\ell = \log(d_\ell^*)$ and allow them to be
        fine-tuned during training. The diversity regulariser
        \cref{eq:diversity_reg} prevents collapse. This combines the
        interpretability of spectral analysis with the flexibility of learnable
        parameters.
\end{enumerate}

This hybrid \textbf{Spectral-Init + Soft-Fine-tune} strategy adds only $L$
extra parameters ($\lambda_1,\ldots,\lambda_L$), is trivially backward-compatible
with the existing codebase, and is expected to converge faster than random
initialisation of the dilation parameters.

For datasets with known non-stationarity, \textbf{Approach II (SKNet)} is
preferred as a drop-in replacement for the dilation stack, at the cost of
$K\times$ memory usage.

%=============================================================================
\section{Implementation Notes}
%=============================================================================

\subsection{Integration with Existing Codebase}

All five approaches operate as drop-in replacements for the
\texttt{DilatedCausalConv} class in
\texttt{models/dilated\_causal\_conv.py}. The interface contract is preserved:

\begin{itemize}
  \item \textbf{Input:} \texttt{x} of shape \texttt{(batch, hidden\_dim, seq\_len)}.
  \item \textbf{Output:} same shape.
  \item \textbf{Config keys to add:}
        \texttt{adaptive\_dilation\_mode} $\in$
        \texttt{\{learnable, sknet, darts, spectral, attention\}}.
\end{itemize}

\subsection{Approach I Implementation Sketch}

\begin{algorithm}[H]
  \SetAlgoLined
  \SetKwInOut{Init}{Initialise}
  \Init{$\lambda_\ell \leftarrow \log(2^{\ell-1})$, $\ell=1,\ldots,L$}
  \KwIn{hidden sequence $\hv^{(0)}$}
  \KwOut{implicit features $\hv^{(L)}$}
  \BlankLine
  \For{$\ell = 1$ \KwTo $L$}{
    $d_\ell \leftarrow \exp(\lambda_\ell)$ \tcp*{continuous dilation}
    $\hv_{\mathrm{conv}} \leftarrow \mathrm{InterpConv}(\hv^{(\ell-1)}, \wv_\ell, d_\ell)$
      \tcp*{interpolated causal conv}
    $\hv^{(\ell)} \leftarrow w_\ell \cdot \mathrm{BN}(\hv_{\mathrm{conv}}) +
      (1-w_\ell) \cdot \hv^{(\ell-1)}$
      \tcp*{residual blend}
  }
  \Return $\hv^{(L)}$
  \caption{Forward pass --- Approach I (Learnable Soft Dilation)}
  \label{alg:approach1}
\end{algorithm}

\subsection{Approach IV Implementation Sketch}

\begin{algorithm}[H]
  \SetAlgoLined
  \SetKwInOut{Init}{Initialise}
  \KwIn{training set $\{\xv^{(n)}\}_{n=1}^N$, sequence length $T$, kernel size $k$, $L$}
  \KwOut{dilation schedule $\{d_1,\ldots,d_L\}$}
  \BlankLine
  $A \leftarrow \mathbf{0}_{\lfloor T/2 \rfloor}$ \tcp*{accumulate spectrum}
  \For{each $\xv^{(n)}$ in training set}{
    $A \mathrel{+}= \frac{1}{C}\sum_c |\FFT(\xv^{(n)}[c,:])|$\;
  }
  $A \leftarrow A / N$\;
  $\{f_1,\ldots,f_L\} \leftarrow \TopK(A[1:], L)$ \tcp*{skip DC}
  \For{$\ell = 1$ \KwTo $L$}{
    $P_\ell \leftarrow \lfloor T / f_\ell \rceil$\;
    $d_\ell \leftarrow \max(1, \lfloor P_\ell / (k-1) \rceil)$\;
  }
  Sort $\{d_\ell\}$ in ascending order\;
  Resolve duplicates via \cref{eq:dedup}\;
  \Return $\{d_1,\ldots,d_L\}$
  \caption{Pre-training spectral dilation calibration --- Approach IV}
  \label{alg:approach4}
\end{algorithm}

%=============================================================================
\section{Conclusion}
%=============================================================================

This document has presented five principled strategies for replacing the fixed
exponential dilation schedule in the SSSDTCN implicit feature extraction module
with adaptive, data-driven alternatives. The key findings are:

\begin{enumerate}
  \item \textbf{Approach I (Learnable Soft Dilation)} is the simplest to implement,
        requiring only $L$ additional scalar parameters, and provides global
        dataset-level adaptation.
  \item \textbf{Approach II (SKNet)} provides the richest per-sample and
        per-channel adaptation with a well-established training procedure, at
        the cost of $K$ parallel branches.
  \item \textbf{Approach III (DARTS)} is the most flexible in exploring discrete
        dilation spaces but requires a two-phase training protocol.
  \item \textbf{Approach IV (Spectral)} requires no learning and can be applied
        immediately; it leverages the known periodicity of industrial processes
        for interpretable, physics-guided dilation.
  \item \textbf{Approach V (Attention-Guided)} is the most expressive,
        providing per-position adaptation, but incurs the highest computational
        overhead.
\end{enumerate}

The recommended starting point is the \textbf{Spectral-Init + Soft-Fine-tune}
hybrid of Approaches IV and I, which combines zero-cost initialisation with
gradient-based refinement and minimal code changes.

Future work should empirically validate each approach on the Debutanizer and
SRU datasets across varying missing ratios, measuring both imputation accuracy
(MAE, RMSE) and the interpretability of the learned dilation patterns.

%=============================================================================
\bibliographystyle{plainnat}
\begin{thebibliography}{99}

\bibitem[Chen et al.(2020a)]{chen2020dynamic}
Chen, Y., Dai, X., Liu, M., Chen, D., Yuan, L., \& Liu, Z. (2020).
\newblock Dynamic convolution: Attention over convolution kernels.
\newblock In \textit{Proceedings of the IEEE/CVF Conference on Computer Vision
  and Pattern Recognition (CVPR)}, pp.\ 11030--11039.

\bibitem[Chen et al.(2020b)]{chen2020temporal}
Chen, T., Liu, S., Chang, S., Cheng, Y., Amini, L., \& Wang, Z. (2020).
\newblock Temporal deformable convolution for action detection.
\newblock \textit{arXiv preprint arXiv:2012.09401}.

\bibitem[Chen et al.(2019)]{chen2019progressive}
Chen, X., Xie, L., Wu, J., \& Tian, Q. (2019).
\newblock Progressive differentiable architecture search: Bridging the depth
  gap between search and evaluation.
\newblock In \textit{Proceedings of the IEEE/CVF International Conference on
  Computer Vision (ICCV)}, pp.\ 1294--1303.

\bibitem[Dai et al.(2017)]{dai2017deformable}
Dai, J., Qi, H., Xiong, Y., Li, Y., Zhang, G., Hu, H., \& Wei, Y. (2017).
\newblock Deformable convolutional networks.
\newblock In \textit{Proceedings of the IEEE International Conference on
  Computer Vision (ICCV)}, pp.\ 764--773.

\bibitem[Gu et al.(2021)]{gu2021efficiently}
Gu, A., Goel, K., \& R\'{e}, C. (2021).
\newblock Efficiently modeling long sequences with structured state spaces.
\newblock In \textit{International Conference on Learning Representations
  (ICLR)}.

\bibitem[Ho et al.(2020)]{ho2020denoising}
Ho, J., Jain, A., \& Abbeel, P. (2020).
\newblock Denoising diffusion probabilistic models.
\newblock In \textit{Advances in Neural Information Processing Systems
  (NeurIPS)}, Vol.\ 33, pp.\ 6840--6851.

\bibitem[Hu et al.(2018)]{hu2018squeeze}
Hu, J., Shen, L., \& Sun, G. (2018).
\newblock Squeeze-and-excitation networks.
\newblock In \textit{Proceedings of the IEEE/CVF Conference on Computer Vision
  and Pattern Recognition (CVPR)}, pp.\ 7132--7141.

\bibitem[Li et al.(2019)]{li2019selective}
Li, X., Wang, W., Hu, X., \& Yang, J. (2019).
\newblock Selective kernel networks.
\newblock In \textit{Proceedings of the IEEE/CVF Conference on Computer Vision
  and Pattern Recognition (CVPR)}, pp.\ 510--519.

\bibitem[Liu et al.(2019)]{liu2019darts}
Liu, H., Simonyan, K., \& Yang, Y. (2019).
\newblock DARTS: Differentiable architecture search.
\newblock In \textit{International Conference on Learning Representations
  (ICLR)}.

\bibitem[SSSDTCN(2024)]{sssdtcn2024}
Anonymous. (2024).
\newblock An implicit-explicit diffusion model for industrial data imputation.
\newblock \textit{Project repository, Yongying-Zhu/sssdtcn}.

\bibitem[Auto-TCN(2022)]{autotcn2022}
Wang, H., Peng, J., Huang, F., Wang, J., Chen, J., \& Xiao, Y. (2022).
\newblock MICN: Multi-scale isometric convolution network for long-term time
  series forecasting.
\newblock In \textit{International Conference on Learning Representations
  (ICLR)}.

\bibitem[TCAN(2020)]{tcan2020}
Pantiskas, L., Verstoep, K., Broekens, J., \& Bal, H. (2020).
\newblock Interpretable multivariate time series forecasting with temporal
  convolutional attention.
\newblock \textit{arXiv preprint arXiv:2011.11638}.

\bibitem[Wang et al.(2018)]{wang2018nonlocal}
Wang, X., Girshick, R., Gupta, A., \& He, K. (2018).
\newblock Non-local means networks.
\newblock In \textit{Proceedings of the IEEE/CVF Conference on Computer Vision
  and Pattern Recognition (CVPR)}, pp.\ 7794--7803.

\bibitem[Wu et al.(2021)]{wu2021autoformer}
Wu, H., Xu, J., Wang, J., \& Long, M. (2021).
\newblock Autoformer: Decomposition transformers with auto-correlation for
  long-term series forecasting.
\newblock In \textit{Advances in Neural Information Processing Systems
  (NeurIPS)}, Vol.\ 34.

\bibitem[Wu et al.(2023)]{wu2023timesnet}
Wu, H., Hu, T., Liu, Y., Zhou, H., Wang, J., \& Long, M. (2023).
\newblock TimesNet: Temporal 2D-variation modeling for general time series
  analysis.
\newblock In \textit{International Conference on Learning Representations
  (ICLR)}.

\bibitem[Zhou et al.(2022)]{zhou2022fedformer}
Zhou, T., Ma, Z., Wen, Q., Wang, X., Sun, L., \& Jin, R. (2022).
\newblock FEDformer: Frequency enhanced decomposed transformer for long-term
  series forecasting.
\newblock In \textit{International Conference on Machine Learning (ICML)},
  pp.\ 27268--27286.

\end{thebibliography}

\end{document}
